---
title: "Economic Shocks and Civil War"
author: "Erin Cavanagh"
description: "Observing the possible relationship between economic shocks to a state's oil market and the likelihood of civil war"
output:
  distill::distill_article:
    self_contained: false
---


## Project thoughts


 
INTRODUCTION

  Civil wars are often the result of commitment problems between the state and a subgroup that has strong motivation to impose their will onto another faction. The cost of a civil war is high to each group involved, and the potential benefit to be gained by the victorious faction must outweigh the costs that will be imposed by the enemy. A possible motive for a group to initiate a civil conflict is the greed and grievance that accompanies economic shocks. If greed was the primary motivation, a portion of the country's population would be incentivized by the potential of control and material gains. Additionally, such greed could be paired a motivation to redress sociopolitical issues that have caused the group grievances (poor labor conditions, high unemployment, etc.) This work aims to discover if countries that suffer from economic shocks are more at risk for civil war outbreaks. More specifically, the work it tests the hypothesis that negative economic shocks to sovereign states leads to the increased likelihood of a civil war conflict in countries. 
  This hypothesis is based in multiple assumptions about the conditions that contribute to the civil tensions. For example, Ii the wake of economic shocks, people may suffer from financial constraints, making the continuation of labor under these circumstances appears less attractive. Further, if a change in production lowers wages in an area, the opportunity cost of joining rebel groups is lowered (so long as the individual believes that there will be a gain in doing so). Additionally, an increase in prices– specifically those of the oil market– can lead to rapacity and violence. For this reason, it seems that states that depend on the exportation of oil for a portion of their GDP will be more sensitive to the effects of economic shocks on internal conflicts. While there are confounding variables (mountainous terrain, varying state capacities, etc.) If correct, the hypothesis would be expect to see a relationship between economic shocks and the outbreak of civil war. 
  In order to support the work's hypothesis, the models should depict a relationship lower per capita incomes and the presence of civil war. Further, it would be expected to see greater conflict in nations that have a larger economic dependence on oil exportation (or percentage of the GDP that comes from oil production). In a regression model, I expect that a high correlation between the income per capita (as well as with Oil and the presence of war in the previous year) and civil war.



Research Question: Are countries that suffer from economic shocks more at risk for civil war outbreaks?

Hypothesis: The grievance of Economic shocks to sovereign states leads to the increased likelihood of a civil war conflict. 




DATA and RESEARCH DESIGN

  This work uses data from the Civil War data set (https://www.dropbox.com/s/gde30hr9yokqku5/civilwar.csv?dl=0).

  This is a observational study, as social scientists are unable to randomize the experiment given the nature of the study. In this case, the occurance of a negative economic shock is considered to be a "naturally assigned" treatment, allowing for social scientists to calculate what the average treatment effect is by comparing countries who undergo economic shocks with those who do not. The observational study is of a cross-sectional design: the outcomes of the treated and control groups are compared at one point in time. In the context of this study, the presence of a civil conflict is compared between countries that have suffering economies and those that do not. This type of research design allows for greater external validity, as a cross-sectional and observational study allows for larger samples that will be more representative of the dynamics in civil conflicts outside of this experiment.

  While the concepts that contributed the forming the hypothesis consider multiple variables, this work looks to simplify such abstractions through its use of fewer variables. This study uses the log of a nation's per capita income as the  independent variable of interest (represented in the code by 'lngdp_'). This is used as an indicator of the health of the country's economy; it demonstrates the general economic production value of each citizen, and gives a general idea of the quality of life in the nation. The plot below filters the data to only compare countries from the Middle East and North America (a necessary adjustment to allow for proper analysis of the impact of a nation's oil exportation later in the experiment). The experiment includes a multiple linear regression model that utilizes two additional independent variables: 'warl' and 'Oil'. 'warl' indicates the lag of a civil war; in other words, it shows data scientists whether the country had a civil conflict in the previous year. 'Oil' is used to decipher which countries have more than 1/3 of their revenue coming from the exportation of Oil. These explanatory variables can be used in conjunction--or studied in their respective relationship to the dependent variable-- to understand the relationship between the state of a nation's economy and the likelihood of civil war. In this research, the outcome variable-- or the dependent variable-- is represented by 'warsa'. This demonstrates the presence of civil conflicts in a nation within a five year period. The "Oil Production and Social Fractionalization in the Middle East and North America" plot also uses 'frac' (representing the social fractionalization in a country) as a dependent variable. 



RESULTS 

  The plot "Log of GDP per capita and Civil War in the Middle East and North America" shows the relationship between the strength of a state's economy and the civil war in Middle Eastern and North American countries. This model uses 'lngdp_' as the independent variable. The log of per capita income is used as it allows data scientists to gain a general understanding of the average economic production value that can be attributed to each citizen. The dependent variable here is the onset of a civil war (in a 5 year period). It appears that countries with low per capita incomes have a much higher likelihood that there will be a civil conflict within the next 5 years. However, the relationship shown in this graph is relatively weak, and is not indicative of a strong causal relationship between the two variables. 

  The boxplot titled "Oil Production and Social Fractionalization in the Middle East and North America" compares the levels of civil fractionalization between countries with economies are highly dependent on oil production/exportation and with those that are not. The use of 'frac' as the dependent variable is meant to simulate the civil tensions that arise prior to (and during) civil disputes. The hypothesis would predict that countries with economic dependence on Oil exportation are more sensitive to civil conflicts, and therefore should have higher internal fractionalization. However, the data from the box plot does not support the hypothesis. In fact, the median fractionalization of the 'Non-Produce' countries is actually higher than that of the 'Producers'. There is a slightly greater spread of the data in the data for 'Producer' countries, while the 'Non-producers' have data that is more concentrated in the lower values of fractionalization.  

  The regression considers the impact of multiple independent variables on the likelihood of civil war in a nation, such as the log of the income per capita, the existence of civil war the previous war, and if Oil accounts for more than 1/3 of the nation's exports. The regression model produces a coefficient of -0.039 for 'lngdp_', with a very small standard error value of 0.008 (representing the absolute value of the average distance any given data point falls from the regression line). This represents a negative relationship between a nation's per capita income and the presence of civil war, albeit a very weak one. While the p value for 'lngdp_' was of a very small value (p=<0.001), the regression model not not show any meaningful relationship between a country's income per capita and the presence of civil war. However, the regression model shows a more promising result for the 'warl' coefficient. With a value of 0.154, the model suggests a positive relationship between the two variables. In the context of this study, this demonstrates the likelihood of a lagging effect of war; the presence of civil war the previous year is correlated to civil war in the cominv years. With a small s.e. and a p= <0.001, this relationship is string than that shown in 'lngdp', but still relatively weak. Finally, the regression model produced a coefficient of 0.068 for 'Oil'. Similar to that of 'lngdp_', the model shows a small s.e. value (0.024) and p value (p=0.006), but still is not indicative of a strong relationship between the degree of Oil exportation by a country and the presence of civil war. Overall, the results of the regression model do not suggest a significant causal relationship between economic factors and the presence of civil war.  




 
```{r}
library(readr)
library(tidyverse)

civilwar <- read_csv("civilwar.csv")
civilwar
cor(civilwar$lngdp_, civilwar$warsa, use = "pairwise")

oil_war_plot <- civilwar|>
  group_by(nafrme)|>
  ggplot(aes(x= lngdp_, y= warsa))+
  geom_smooth(method ="loess", se = TRUE, size= 1.1)+
  labs(title = "Log of GDP per capita and Civil War in the Middle East and North America", x= "Log per capita Income", y ="Onset of Civil War (in 5 year period)")

oil_war_plot
   
treat_means <- civilwar|>
filter(warsa==1)|>
summarize(trtgdp_mean = mean(lngdp_, na.rm = TRUE))
treat_means

control_means <- civilwar|>
  filter(warsa==0)|>
  summarize( congdpmean= mean(lngdp_, na.rm = TRUE))
control_means

 ate <- treat_means - control_means
  ate

```




 

```{r}

civilwar <- civilwar|>
  mutate(
    oil_status = if_else(Oil ==1, "Producer", "Non-Producer")
  )

civilwar


  warplot <- civilwar|>
group_by(nafrme)|>
ggplot(civilwar, mapping= aes(x=oil_status, y=frac))+
  geom_boxplot()+
  labs(title= "Oil Production and Social Fractionalization in the Middle East and North America", x= "Country as an High Producer of Oil (accounts for >1/3 of its exports)", y = "Social Fractionalization")
warplot
```


```{r}

  econ_effect <- lm(warsa ~ lngdp_ + warl+ Oil, data = civilwar)
econ_effect

modelsummary::modelsummary(econ_effect,
                           statistic = c("s.e. = {std.error}", 
                                         "p = {p.value}"),
                           gof_map = c("nobs", "r.squared", "adj.r.squared"))   
```
     

CONCLUSION

  According to political theory, civil wars are often the result of collective grievances that motivate a group to rebel. This work was intended to discover any possible relationships between economic factors and the onset of civil wars. This was done by isolating variables that would represent the general health of a country's economy (such as the income per capita) as well as other factors that would make a country more susceptible to internal conflict (ie. economic dependence on the Oil market, conflict in previous years, and fractionalization). These variables were manipulated to cross-sectionally observe their respective effects on civil conflicts. However, the models produced showed limited support for the hypothesis. While there were small relationships shown, none offered proof of a significant causal relationship between these economic factors and civil war in countries in the Middle East and Northern America. This could be due to confounding variables that were not accounted for (ie. structure of the government in power, mountainous terrain, ideological beliefs, etc.). Additionally, the study could be adjusted to address issues that led to the failed hypothesis. For example, the sample could select different countries in future attempts, comparing nations with similar terrain, population sizes, exports, government structures, etc. to limit the effect of confounding variables. 


